---
title: "The Evolution of Human Eating"
format: html
editor: visual
---

```{r setup, include=FALSE}
# Load necessary packages for this chapter
library(tidyverse)
library(knitr)
library(ggpubr)
library(patchwork)
library(here)

# Set default options
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 300,
  out.width = "100%"
)

# Set consistent theme for all visualizations
theme_set(theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "grey70", fill = NA, size = 0.5),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 12, color = "grey50")
  )
)
```

# Early Hominid Dietary Patterns

" QOUTE ASCENT OF MAN"---

The story of human diet is etched into the very structure of our teeth and jaws. As we examine the fossil record, a remarkable narrative emerges—one that connects our biological adaptations to changing food resources over millions of years.

There is a popular impression, often reflected by fad diets -paelo etc, that the our ancestors diet was "healthier" when coimpared to the modern industrial based foods consumed today. There are two major problems with this simplistic view. FIrstly, we do not have an accurate record of the precise foods, drinks, diet, etc -more food vessels than foods AND the idea of a hunter/gatherer as the main soruce of food is based on tenuous evidence. EPXAND

(MOVE TO LATER) There are biases in our understadning of the food industry and intereationshuips are complex- logistics, production mass, agricult practices, food industry/advertising, safe practiuces, water, health, .....wealth...ethnic, cultural,

## Dental Evidence Through Time

The teeth of our earliest ancestors tell a compelling story of dietary adaptation. Australopithecines, living 4-2 million years ago, possessed dentition markedly different from modern humans. Their molars were larger, with thicker enamel, adapted for processing tough, fibrous plant materials that required substantial grinding force [@ungar2012dental].

```{r tooth-comparison, fig.cap="Comparison of molar size and shape across human evolution, showing the reduction in size and complexity over time."}
# Sample code to create evolutionary tooth comparison
# In actual use, you would import real data

# Create sample data for visualization
hominid_teeth <- tibble(
  Species = factor(c("Australopithecus", "Early Homo", "Neanderthal", "Modern Human"),
                  levels = c("Australopithecus", "Early Homo", "Neanderthal", "Modern Human")),
  Molar_Size = c(1.8, 1.5, 1.2, 1.0),  # Relative sizes
  Enamel_Thickness = c(2.0, 1.7, 1.3, 1.0),  # Relative thickness
  Complexity = c(2.0, 1.8, 1.3, 1.0)  # Relative complexity
)

# Create visualization
ggplot(hominid_teeth, aes(x = Species, y = Molar_Size)) +
  geom_bar(stat = "identity", fill = "skyblue", alpha = 0.7) +
  geom_point(aes(y = Enamel_Thickness), color = "darkred", size = 3) +
  geom_line(aes(y = Enamel_Thickness, group = 1), color = "darkred") +
  labs(
    title = "Evolution of Human Molar Characteristics",
    subtitle = "Showing reduction in size and enamel thickness over time",
    y = "Relative Value (Modern Human = 1)",
    x = NULL
  ) +
  scale_y_continuous(limits = c(0, 2.2))
```

Microscopic analysis of tooth wear patterns reveals crucial information about diet composition. While the large molars of Australopithecus might suggest a diet of hard objects, microwear analysis of *Australopithecus africanus* from Sterkfontein caves shows patterns more consistent with the consumption of softer, tougher foods like leaves and fruits [@scott2005dental]. This apparent contradiction highlights the complexity of early dietary adaptations.

## Isotope Analysis and Dietary Composition

Carbon isotope ratios (δ^13^C) from fossil remains provide direct evidence of dietary composition. Analysis of Australopithecine remains from South African sites reveals a mixed C~3~/C~4~ signal, indicating consumption of both forest resources (fruits, nuts) and savanna resources (grasses, sedges, or animals that consumed these plants) [@sponheimer2013isotopic].

```{r isotope-analysis, fig.cap="Carbon isotope values (δ¹³C) from fossil hominid samples showing dietary shifts over time."}
# Sample visualization of isotope data
# In actual use, you would import real data

# Create sample isotope data
isotope_data <- tibble(
  Era = c(rep("Early Australopithecine", 8), 
          rep("Late Australopithecine", 10),
          rep("Early Homo", 12), 
          rep("Homo erectus", 8),
          rep("Modern Hunter-Gatherer", 15)),
  d13C = c(rnorm(8, -14, 1), 
           rnorm(10, -12, 1.5),
           rnorm(12, -11, 1.8), 
           rnorm(8, -10, 2),
           rnorm(15, -12, 3)),
  Diet_Type = case_when(
    d13C < -16 ~ "Pure C3 (Forest)",
    d13C > -8 ~ "Pure C4 (Savanna)",
    TRUE ~ "Mixed C3/C4"
  )
)

# Create visualization
ggplot(isotope_data, aes(x = Era, y = d13C, fill = Era)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  geom_hline(yintercept = c(-16, -8), linetype = "dashed", color = "darkgrey") +
  annotate("text", x = 1, y = -17, label = "C3 (Forest Foods)", hjust = 0) +
  annotate("text", x = 1, y = -7, label = "C4 (Savanna Foods)", hjust = 0) +
  labs(
    title = "Carbon Isotope Ratios in Hominid Fossils",
    subtitle = "Showing dietary diversification over time",
    y = "δ¹³C Value (‰)",
    x = NULL
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

This diversification of diet was a crucial adaptation that allowed early hominids to exploit multiple ecological niches, particularly important during periods of climate variability.

## Comparative Analysis with Modern Hunter-Gatherers

The !Kung San people of the Kalahari provide valuable insights into pre-agricultural human diets. Their traditional diet includes over 100 plant species and numerous animal foods, varying seasonally [@lee1979kung]. Dental wear patterns in older !Kung individuals show similarities to certain fossil specimens, suggesting parallels in food processing techniques.

While we must be cautious about direct comparisons between modern groups and our ancestors, such studies offer a window into potential dietary patterns before agriculture.

```{r hunter-gatherer-comparison, fig.cap="Comparison of seasonal food resource use in modern hunter-gatherer societies."}
# Sample data for hunter-gatherer seasonal diet composition
# In actual use, you would import real data

# Create sample seasonal data
seasons <- c("Spring", "Summer", "Fall", "Winter")
hadza_data <- tibble(
  Season = rep(seasons, each = 4),
  Food_Category = rep(c("Tubers", "Fruits/Berries", "Honey", "Animal Foods"), 4),
  Percentage = c(
    40, 25, 5, 30,  # Spring
    30, 45, 10, 15, # Summer
    35, 30, 5, 30,  # Fall
    50, 10, 0, 40   # Winter
  )
)

# Create visualization
ggplot(hadza_data, aes(x = Season, y = Percentage, fill = Food_Category)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Seasonal Diet Composition in Hadza Hunter-Gatherers",
    subtitle = "Showing adaptation to seasonal resource availability",
    y = "Percentage of Diet",
    x = NULL,
    fill = "Food Category"
  ) +
  theme(legend.position = "bottom")
```

The dietary flexibility observed in modern hunter-gatherers likely mirrors adaptations that allowed early humans to survive through seasonal and climate-driven changes in food availability.

# The Revolutionary Impact of Fire

The controlled use of fire represents one of the most transformative innovations in human dietary evolution. While pinpointing the exact timing remains challenging, evidence suggests that *Homo erectus* may have controlled fire by approximately 1 million years ago, with more definitive evidence by 400,000 years ago [@wrangham2009catching].

## Cooking and Nutrient Bioavailability

Cooking fundamentally alters the digestibility and nutritional value of foods, particularly starchy tubers and meat, in ways that would have dramatically improved caloric and nutrient intake for early humans [@carmody2011energetic].

```{r cooking-impact, fig.cap="Impact of cooking on nutrient bioavailability for different food types."}
# Sample data for cooking impact
# In actual use, you would import real data

cooking_impact <- tibble(
  Food_Type = rep(c("Starchy Tubers", "Meat", "Vegetables", "Grains"), each = 2),
  Cooking_Status = rep(c("Raw", "Cooked"), 4),
  Energy_Available = c(
    30, 80,  # Tubers raw vs cooked
    50, 75,  # Meat raw vs cooked
    40, 60,  # Vegetables raw vs cooked
    25, 70   # Grains raw vs cooked
  ),
  Prep_Time = c(
    5, 30,  # Tubers
    10, 40, # Meat
    5, 20,  # Vegetables
    5, 45   # Grains
  )
)

# Create visualization
cooking_plot <- ggplot(cooking_impact, aes(x = Food_Type, y = Energy_Available, 
                                fill = Cooking_Status, group = Cooking_Status)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Raw" = "#E69F00", "Cooked" = "#D55E00")) +
  labs(
    title = "Impact of Cooking on Energy Availability",
    y = "Energy Available (%)",
    x = NULL
  )

time_plot <- ggplot(cooking_impact, aes(x = Food_Type, y = Prep_Time, 
                              fill = Cooking_Status, group = Cooking_Status)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Raw" = "#E69F00", "Cooked" = "#D55E00")) +
  labs(
    title = "Processing Time Requirements",
    y = "Time (minutes)",
    x = NULL
  )

# Combine plots
cooking_plot + time_plot + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

This dramatic increase in energy availability, particularly from starchy foods, likely fueled the expansion of hominin brain size, which requires substantial energetic investment [@leonard2007evolution].

## Dental and Jaw Structure Changes

The adoption of cooking correlates with significant changes in dental and jaw morphology in the *Homo* lineage. Compared to earlier hominids, later *Homo* species show:

-   Reduced molar size
-   Decreased mandibular robusticity
-   Less pronounced muscle attachment sites for chewing muscles
-   Overall facial shortening

These changes reflect reduced mechanical demands for processing food, as cooking softens tough foods and neutralizes plant defense compounds [@zink2014impact].

```{r jaw-changes, fig.cap="Changes in jaw robusticity across human evolution, correlated with cooking adoption."}
# Sample data for jaw characteristics
# In actual use, you would import real data

jaw_data <- tibble(
  Species = c("Australopithecus", "Homo habilis", "Homo erectus", "Homo neanderthalensis", "Homo sapiens"),
  Era = c("4-2 mya", "2.4-1.5 mya", "1.9-0.4 mya", "0.4-0.04 mya", "0.2 mya-present"),
  Jaw_Size = c(95, 85, 70, 65, 55),
  Evidence_of_Fire = c("None", "Contested", "Some", "Abundant", "Universal"),
  Fire_Score = c(0, 0.3, 0.7, 0.9, 1.0)
)

# Create visualization
ggplot(jaw_data, aes(x = Era, y = Jaw_Size, group = 1)) +
  geom_line(size = 1, color = "darkblue") +
  geom_point(aes(size = Fire_Score, color = Fire_Score), alpha = 0.7) +
  scale_color_gradient(low = "#FEE08B", high = "#E6550D") +
  scale_size_continuous(range = c(3, 10)) +
  geom_text(aes(label = Species), vjust = -1.5, size = 3.5) +
  labs(
    title = "Jaw Robusticity and Cooking Evidence Through Human Evolution",
    subtitle = "Showing inverse relationship between fire use and jaw size",
    y = "Relative Jaw Robusticity Index",
    x = NULL,
    color = "Evidence of Fire Use",
    size = "Evidence of Fire Use"
  ) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Oral Health Implications

The transition to cooked foods had complex effects on oral health. While softened foods reduced extreme wear patterns, the increased consumption of starchy foods may have promoted the growth of cariogenic oral bacteria. However, evidence of dental caries remains relatively rare in fossil remains until the agricultural revolution [@humphrey2014evolutionary].

# Case Study: The Qesem Cave Evidence

Qesem Cave in Israel (400,000-200,000 years ago) provides compelling evidence for regular cooking practices by Middle Pleistocene hominins. Microscopic analysis of dental calculus from teeth found at the site revealed:

-   Starch granules with morphological changes consistent with cooking
-   Presence of cooked plant tissues
-   Evidence of controlled use of fire in a defined hearth area

This site demonstrates that by this period, cooking had become a regular, established practice with direct impacts on dental health and diet [@hardy2016dental].

## Technological Innovations

Beyond fire itself, archaeological evidence points to the development of specialized tools for food processing. Ground stone tools appear in the archaeological record by 30,000 years ago, allowing for more efficient processing of plant foods [@fuller2018archaeological]. These technologies would have further reduced the mechanical demands on teeth and jaws.

```{r}
# End of chapter code - could include data export or other processing
# not shown in the rendered document

# Example of saving a dataset for use in a later chapter
# save(hominid_teeth, file = here("data", "processed", "hominid_teeth.RData"))
```

## References {.unnumbered}

<!-- References will be automatically generated from your citations -->

## Early Hominid Dietary Patterns

The teeth of our earliest ancestors tell a compelling story of dietary adaptation. Australopithecines, living 4-2 million years ago, possessed dentition markedly different from modern humans.

### Dental Evidence Through Time

Their molars were larger, with thicker enamel, adapted for processing tough, fibrous plant materials that required substantial grinding force.

### Isotope Analysis and Dietary Composition

Carbon isotope ratios (δ^13^C) from fossil remains provide direct evidence of dietary composition.

## The Revolutionary Impact of Fire

## The controlled use of fire represents one of the most transformative innovations in human dietary evolution.
